Heuristics and Really Hard Instances for Subgraph Isomorphism Problems

### 2016-02-02 paper/paper.tex
James's changes

Throughout, our graphs are unlabelled, undirected, and do not have any loops (vertices which are
adjacent to themselves).  The \emph{order} of a graph is the cardinality of its vertex set. We write
$\operatorname{V}(G)$ for the vertex set of a graph $G$. The \emph{complement} of a graph $G$,
denoted $\shortoverline{G}$, is the
graph with the same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ if and
only if
$v$ and $w$ are not adjacent in $G$. We write $G(v, p)$ for an Erd\H{o}s-R\'enyi random graph with
$v$ vertices, and an edge between each distinct pair of vertices with independent probability $p$.

@@

\begin{figure}[tb]
    \setlength{\abovecaptionskip}{6pt}
    \input{gen-graph-phase-transition.tex}
    \caption{With a fixed a pattern graph order of 20, a target graph order of 150, a target edge
        probability of 0.4, and varying pattern edge probability, we observe a phase transition and
        complexity peak with the Glasgow solver in the non-induced variant. Each point represents
        one instance. The lines show mean search effort, and mean proportion satisfiable.}
    \label{figure:phase-transition}
\end{figure}

@@

\begin{figure}[tb]
    \setlength{\abovecaptionskip}{0pt}
    \input{gen-graph-non-induced.tex}
    \caption{Behaviour of algorithms on the non-induced variant. For each plot, the x-axis is the
        pattern edge probability and the y-axis is the target edge probability, both from 0 to 1.
        Along the top row, we show the proportion of instances which are satisfiable; the white
        bands shows the phase transitions, and the black lines are our predictions of where the
        phase transition will occur. On the second row, we show the number of search nodes used by the
        Glasgow algorithm, on the third row, the number of search nodes used by the LAD algorithm, and the
        fourth, VF2; the dark regions indicate ``really hard'' instances.}
    \label{figure:non-induced}
\end{figure}

What if we alter the edge probabilities for both the pattern graph and the target graph?  In the top
row of \cref{figure:non-induced} we show the satisfiability phase transition for the non-induced
variant, for patterns of order 10, 20 and 30, a target of order 150, and varying pattern (x-axis)
and target (y-axis) edge probabilities. Each axis runs over 51 edge probabilities, from 0 to 1 in
steps of 0.05; for each of these 2601 points, we generate ten random instances. The colour denotes
the proportion of these instances which were found to be satisfiable.  Inside the orange region, at
the bottom right of each plot, every instance is unsatisfiable---here we are trying to find a dense
pattern in a sparse target. In the purple region, at the top left, every instance is
satisfiable---we are looking for a sparse pattern in a dense target (which is easy, since we only
have to preserve adjacency, not non-adjacency). The white band between the regions shows the
location of the phase transition: here, roughly half the instances are satisfiable. (We discuss the
black line below.)

@@

When designing variable and value ordering heuristics for backtracking search algorithms, there are
various general principles which appear to be effective---one of these is to try to maximise the
expected number of solutions inside any subproblem considered during search \citep{Gent:1996:EN}.
This is usually done by cheaper surrogates, rather than direct calculation. When branching, both LAD
and Glasgow pick a variable with fewest remaining values in its domain: doing this will generally
reduce the first part of the $\langle Sol \rangle$ equation by as little as possible. When two or
more domains are of equal size, LAD simply breaks ties lexicographically, whereas Glasgow will pick
a variable corresponding to a pattern vertex of highest degree. This strategy was apparently
determined empirically, but could have been derived from the $\langle Sol \rangle$ formula: picking
a pattern vertex of high degree will make the remaining pattern subgraph sparser, which will
decrease the exponent in the second half of the formula, maximising the overall value. LAD does not
apply a value ordering heuristic, but Glasgow does: it prefers target vertices of lowest degree.
Again, this appears to have been determined empirically, but it has the effect of increasing
$\langle Sol \rangle$ by increasing the remaining target density. The VF2 heuristics, in contrast,
are based around preserving connectivity, which gives very little discrimination except on the
sparsest of inputs.

@@

For patterns of order 10, it is always better to try to move towards the satisfiable region: if we
are in the bottom right diagonal half, we are best retaining the original heuristics (which move us
towards the top right), and if we are in the top left we should use the complement instead. This
goes against a suggestion by \citet{Walsh:1998} that switching heuristics based upon an estimate of
the solubility of the problem may offer good performance.

@@

The region in the parameter space where both pattern and target have medium density is far from a
phase transition, but nevertheless contains instances that are hard for all three solvers. We would
like to know whether this is due to a weakness in current solvers (perhaps our solvers cannot region
about adjacency and non-adjacency simultaneously?), or whether instances in this region are
inherently difficult to solve.  Thus we repeat the induced experiments on smaller pattern and target
graphs, using different solving techniques.  Although these techniques are not competitive in
absolute terms, we wish to see if the same pattern of behaviour occurs. The results are plotted in
\cref{figure:alt}.

@@

The pseudo-boolean (PB) encoding is as follows. For each pattern vertex $v$ and each target
vertex $w$, we have a binary variable which takes the value 1 if and only if that $v$ is mapped to
$w$.  Constraints are added to ensure that each pattern vertex maps to exactly one target vertex,
that each target vertex is mapped to by at most one pattern vertex, that adjacent vertices are
mapped to adjacent vertices, and that non-adjacent vertices are mapped to non-adjacent vertices. We
used the Clasp solver~\citep{gekakaosscsc11a} version 3.1.3 to solve the pseudo-boolean instances.
The instances that are hard for the Glasgow solver remain hard for the PB solver, including
instances inside the central region, and the easy satisfiable instances remain easy.

>>>>

Throughout, our graphs are unlabelled, undirected, and do not have any loops (vertices which are
adjacent to themselves).  The \emph{order} of a graph is the cardinality of its vertex set. We write
$\operatorname{V}(G)$ for the vertex set of a graph $G$. The \emph{complement} of a graph $G$,
denoted $\shortoverline{G}$, is the
graph with the same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ if and
only if
$v$ and $w$ are not adjacent in $G$. We write $G(n, p)$ for an Erd\H{o}s-R\'enyi random graph with
$n$ vertices, and an edge between each distinct pair of vertices with independent probability $p$.

@@

\begin{figure}[tb]
    \setlength{\abovecaptionskip}{6pt}
    \input{gen-graph-phase-transition.tex}
    \caption{With a fixed a pattern graph order of 20, a target graph order of 150, a target edge
        probability of 0.40, and varying pattern edge probability, we observe a phase transition and
        complexity peak with the Glasgow solver in the non-induced variant. Each point represents
        one instance. The lines show mean search effort and mean proportion satisfiable.}
    \label{figure:phase-transition}
\end{figure}

@@

\begin{figure}[tb]
    \setlength{\abovecaptionskip}{0pt}
    \input{gen-graph-non-induced.tex}
    \caption{Behaviour of algorithms on the non-induced variant. For each plot, the x-axis is the
        pattern edge probability and the y-axis is the target edge probability, both from 0 to 1.
        Along the top row, we show the proportion of instances which are satisfiable; the white
        bands shows the phase transitions, and the black lines are our predictions of where the
        phase transition will occur. On the final three rows, we show the number of search nodes used by the
        Glasgow, LAD and VF2 solvers; the dark regions indicate ``really hard'' instances.}
    \label{figure:non-induced}
\end{figure}

What if we alter the edge probabilities for both the pattern graph and the target graph?  In the top
row of \cref{figure:non-induced} we show the satisfiability phase transition for the non-induced
variant, for patterns of order 10, 20 and 30, targets of order 150, and varying pattern (x-axis)
and target (y-axis) edge probabilities. Each axis runs over 51 edge probabilities, from 0 to 1 in
steps of 0.02; for each of these 2601 points, we generate ten random instances. The colour denotes
the proportion of these instances which were found to be satisfiable.  Inside the orange region, at
the bottom right of each plot, every instance is unsatisfiable---here we are trying to find a dense
pattern in a sparse target. In the purple region, at the top left, every instance is
satisfiable---we are looking for a sparse pattern in a dense target (which is easy, since we only
have to preserve adjacency, not non-adjacency). The white band between the regions shows the
location of the phase transition: here, roughly half the instances are satisfiable. (We discuss the
black line below.)

@@

Various general principles have been considered when designing variable and value ordering
heuristics for backtracking search algorithms---one of these is to try to maximise the expected
number of solutions inside any subproblem considered during search \citep{Gent:1996:EN}.  This is
usually done by cheaper surrogates, rather than direct calculation. When branching, both LAD and
Glasgow pick a variable with fewest remaining values in its domain: doing this will generally reduce
the first part of the $\langle Sol \rangle$ equation by as little as possible. When two or more
domains are of equal size, LAD simply breaks ties lexicographically, whereas Glasgow will pick a
variable corresponding to a pattern vertex of highest degree. This strategy was apparently
determined empirically, but could have been derived from the $\langle Sol \rangle$ formula: picking
a pattern vertex of high degree will make the remaining pattern subgraph sparser, which will
decrease the exponent in the second half of the formula, maximising the overall value. LAD does not
apply a value ordering heuristic, but Glasgow does: it prefers target vertices of lowest degree.
Again, this appears to have been determined empirically, but it has the effect of increasing
$\langle Sol \rangle$ by increasing the remaining target density. The VF2 heuristics, in contrast,
are based around preserving connectivity, which gives very little discrimination except on the
sparsest of inputs.

@@

For patterns of order 10, it is always better to try to move towards the satisfiable region: if we
are in the bottom right diagonal half, we are best retaining the original heuristics (which move us
towards the top left), and if we are in the top left we should use the complement instead. This
goes against a suggestion by \citet{Walsh:1998} that switching heuristics based upon an estimate of
the solubility of the problem may offer good performance.

@@

The region in the parameter space where both pattern and target have medium density is far from a
phase transition, but nevertheless contains instances that are hard for all three solvers. We would
like to know whether this is due to a weakness in current solvers (perhaps our solvers cannot reason
about adjacency and non-adjacency simultaneously?), or whether instances in this region are
inherently difficult to solve.  Thus we repeat the induced experiments on smaller pattern and target
graphs, using different solving techniques.  Although these techniques are not competitive in
absolute terms, we wish to see if the same pattern of behaviour occurs. The results are plotted in
\cref{figure:alt}.

@@

The pseudo-boolean (PB) encoding is as follows. For each pattern vertex $v$ and each target
vertex $w$, we have a binary variable which takes the value 1 if and only if $v$ is mapped to
$w$.  Constraints are added to ensure that each pattern vertex maps to exactly one target vertex,
that each target vertex is mapped to by at most one pattern vertex, that adjacent vertices are
mapped to adjacent vertices, and that non-adjacent vertices are mapped to non-adjacent vertices. We
used the Clasp solver~\citep{gekakaosscsc11a} version 3.1.3 to solve the pseudo-boolean instances.
The instances that are hard for the Glasgow solver remain hard for the PB solver, including
instances inside the central region, and the easy satisfiable instances remain easy.
